NumLin: Linear Types for Linear Algebra

### 2018-11-05 write-up/paper/appendix.tex
Add section on OCaml 'optimisation'
Add frac. perm. sub. lemma
Start on Formal System
Add Ott Semantics
Add interpretation and proof
Start on Soundness Theorem

\section{Appendix}

\begin{figure}[t]
\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    x[e] &
    \Rightarrow &
    \mathbf{get}\ \_\ x\ (e) \;\qquad \textrm{(similarly for matrices)}
\\
    x[e_1] := e_2 &
    \Rightarrow &
    \mathbf{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
\\
\\
    pat & ::= & ()\ \mid\ x\ \mid\ !x\ \mid\ \mathbf{Many\ } pat\ \mid\ (pat, pat)
\\
    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{\mathbf{let\ Many}\ x = e_1\ \mathbf{in\ } \\
    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2}
\\
    \mathbf{let\ Many} \langle pat_x \rangle\ = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \mathbf{let\ Many}\ x = x\ \mathbf{in\ } \\
        \mathbf{let\ }\ \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
\\
    \mathbf{let}\ (\langle pat_a \rangle, \langle pat_b \rangle)\  = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \mathbf{let\ }\ (a,b)\ = a\_b\ \mathbf{in\ }
        \ \mathbf{let\ }\ \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
        \mathbf{let\ }\ \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
\\
    \mathbf{fun}\ (\langle pat_x \rangle\ : t) \rightarrow e &
    \Rightarrow &
    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ }\ \langle pat_x \rangle = x\ \mathbf{in\ } e
\\
\\
    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
\\
    \mathbf{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
    \Rightarrow &
    \mathbf{fun}\ \langle arg_1 \rangle \rightarrow ..
    \ \mathbf{fun}\ \langle arg_n \rangle \rightarrow e
\\
    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \mathbf{in}\ e_2
\\
    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \mathbf{in}\ e_2
\\
    \mathbf{let}\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle arg_{1 .. n} \rangle}
    \rightarrow e_1)\ \mathbf{in}\ e_2
\\
    \mathbf{let\ rec}\ f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{\mathrm{fixpoint} \equiv \mathbf{fix}\ (f, x : t,
        \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} ) \\
    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2}
    \end{array}
\]
\end{center}
\caption{Desugaring from \lang\ concrete syntax to core constructs.}\label{fig:lang_desugar}
\end{figure}

\begin{figure}[p]
    \centering
    \begin{minted}[fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let sigma_h = Prim.matrix k n in
  let (sigma, h), sigma_h =
    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) sigma_h
  in
  let (sigma_h, h), r_2 =
    Prim.gemm (Many 1.) (sigma_h, Many false) (h, Many true) (Many 1.) r_1
  in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
  in
  let h, new_h = Prim.copy_mat_to h sigma_h in
  let r_2, new_r = Prim.copy_mat r_2 in
  let chol_r, sol_h = Prim.posv new_r new_h in
  let chol_r, sol_data = Prim.potrs chol_r data_2 in
  let () = Prim.free_mat chol_r in
  let h_sol_h = Prim.matrix n n in
  let (h, sol_h), h_sol_h =
    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
  in
  let () = Prim.free_mat sol_h in
  let h_sol_data = Prim.matrix n (Many 1) in
  let (h, sol_data), h_sol_data =
    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
  in
  let mu, mu_copy = Prim.copy_mat mu in
  let (sigma, h_sol_data), new_mu =
    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
  in
  let () = Prim.free_mat h_sol_data in
  let h_sol_h_sigma = Prim.matrix n n in
  let (sigma, h_sol_h), h_sol_h_sigma =
    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
  in
  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
  let (sigma, h_sol_h_sigma), new_sigma =
    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
  in
  let () = Prim.free_mat h_sol_h_sigma in
  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
in
kalman
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated (at \emph{compile time})
        from the code in Figure~\ref{fig:lang_kalman}, passed through
        \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}

\end{figure}

>>>>

\section{\lang\ Specification}

\ottstyledefaults{premiselayout=justify}%
\subsection{Static Semantics}\label{subsec:static_sem}
\ottdefnsTypes%

\subsection{Dynamic Semantics}\label{subsec:dyn_sem}
\ottdefnsOpXXSem%

\input{../semantics/interpretation}

\input{../semantics/lemmas}

\input{../semantics/soundness}

\subsection{Well-formed types}
\ottdefnsWellXXFormed%

\section{\lang\ Grammar}\label{sec:grammar_def}
\ottgrammar%

\section{Desugaring \lang}\label{sec:lang_desugar}

\begin{center}
\[\def\arraystretch{1.3}
    \begin{array}{rcl}
    x[e] &
    \Rightarrow &
    \mathbf{get}\ \_\ x\ (e) \;\qquad \textrm{(similarly for matrices)}
\\
    x[e_1] := e_2 &
    \Rightarrow &
    \mathbf{set}\ x\ (e_1)\ (e_2) \quad \textrm{(similarly for matrices)}
\\
\\
    pat & ::= & ()\ \mid\ x\ \mid\ !x\ \mid\ \mathbf{Many\ } pat\ \mid\ (pat, pat)
\\
    \mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{\mathbf{let\ Many}\ x = e_1\ \mathbf{in\ } \\
    \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\ \mathbf{in}\ e_2}
\\
    \mathbf{let\ Many} \langle pat_x \rangle\ = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \mathbf{let\ Many}\ x = x\ \mathbf{in\ } \\
        \mathbf{let\ } \langle pat_x \rangle\ = x\ \mathbf{in\ } e_2}
\\
    \mathbf{let}\ (\langle pat_a \rangle, \langle pat_b \rangle)\  = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \specialcell[t]{l}{%
        \mathbf{let\ } (a,b)\ = a\_b\ \mathbf{in\ }
        \ \mathbf{let\ } \langle pat_a \rangle\ = a\ \mathbf{in\ } \\
        \mathbf{let\ } \langle pat_b \rangle\ = b\ \mathbf{in\ } e_2}
\\
    \mathbf{fun}\ (\langle pat_x \rangle : t) \rightarrow e &
    \Rightarrow &
    \mathbf{fun}\ (x : t) \rightarrow \mathbf{let\ } \langle pat_x \rangle = x\ \mathbf{in\ } e
\\
\\
    arg & ::= & \langle pat \rangle : t\ \mid\ {'x} \textrm{ (fractional permission variable)}
\\
    \mathbf{fun}\ \langle arg_{1 .. n} \rangle \rightarrow e &
    \Rightarrow &
    \mathbf{fun}\ \langle arg_1 \rangle \rightarrow ..
    \ \mathbf{fun}\ \langle arg_n \rangle \rightarrow e
\\
    \mathbf{let}\ f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let}\ f = \mathbf{fun}\ {\langle arg_{1 .. n} \rangle} \rightarrow e_1\
    \mathbf{in}\ e_2
\\
    \mathbf{let}\ !f\ {\langle arg_{1 .. n} \rangle} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let\ Many}\ f = \mathbf{Many}\ (\mathbf{fun}\ {\langle arg_{1 .. n} \rangle}
    \rightarrow e_1)\ \mathbf{in}\ e_2
\\
    \mathrm{fixpoint} & \equiv & \mathbf{fix}\ (f, x : t, \mathbf{fun}
    \ {\langle arg_{1 .. n} \rangle} \rightarrow e_1 : {t'} )
\\
    \mathbf{let\ rec}\ f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let}\ f = \mathrm{fixpoint}\ \mathbf{in}\ e_2
\\
    \mathbf{let\ rec}\ !f\ (x : t)\ {\langle arg_{1 .. n} \rangle} : {t'} = e_1\ \mathbf{in}\ e_2 &
    \Rightarrow &
    \mathbf{let\ Many}\ f = \mathbf{Many}\ \mathrm{fixpoint}\ \mathbf{in}\ e_2
    \end{array}
\]
\end{center}

\clearpage
\section{Primitives}\label{subsec:primitives}

\vspace*{\fill}
\begin{center}
\begin{minted}[fontsize=\small]{ocaml}
module Arr = Owl.Dense.Ndarray.D
type z = Z
type 'a s = Succ
type 'a arr = A of Arr.arr [@@unboxed]
type 'a mat = M of Arr.arr [@@unboxed]
type 'a bang = Many of 'a [@@unboxed]
module Prim :
sig
  val extract : 'a bang -> 'a
  (** Boolean *)
  val not_ : bool bang -> bool bang
  (** Arithmetic, many omitted for brevity *)
  val addI : int bang -> int bang -> int bang
  val eqI : int bang -> int bang -> bool bang
  (** Arrays *)
  val set : z arr -> int bang -> float bang -> z arr
  val get : 'a arr -> int bang -> 'a arr * float bang
  val share : 'a arr -> 'a s arr * 'a s arr
  val unshare : 'a s arr -> 'a s arr -> 'a arr
  val free : z arr -> unit
  (** Owl *)
  val array : int bang -> z arr
  val copy : 'a arr -> 'a arr * z arr
  val sin : z arr -> z arr
  val hypot : z arr -> 'a arr -> 'a arr * z arr
  (** Level 1 BLAS *)
  val asum : 'a arr -> 'a arr * float bang
  val axpy : float bang -> 'a arr -> z arr -> 'a arr * z arr
  val dot : 'a arr -> 'b arr -> ('a arr * 'b arr) * float bang
  val scal : float bang -> z arr -> z arr
  val amax : 'a arr -> 'a arr * int bang
  (* Matrix, some omitted for brevity *)
  val matrix : int bang -> int bang -> z mat
  val eye : int bang -> z mat
  val copy_mat : 'a mat -> 'a mat * z mat
  val copy_mat_to : 'a mat -> z mat -> 'a mat * z mat
  val size_mat : 'a mat -> 'a mat * (int bang * int bang)
  val transpose : 'a mat -> 'a mat * z mat
  (* Level 3 BLAS/LAPACK *)
  val gemm : float bang -> ('a mat * bool bang) -> ('b mat * bool bang) ->
             float bang -> z mat -> ('a mat * 'b mat) * z mat
  val symm : bool bang -> float bang -> 'a mat -> 'b mat -> float bang ->
             z mat -> ('a mat * 'b mat) * z mat
  val gesv : z mat -> z mat -> z mat * z mat
  val posv : z mat -> z mat -> z mat * z mat
  val potrs : 'a mat -> z mat -> 'a mat * z mat
  val syrk : bool bang -> float bang -> 'a mat -> float bang -> z mat ->
           'a mat * z mat
end
\end{minted}
\end{center}
\vfill

\clearpage
\section{Kalman Filters from \lang\ and C}

\vspace*{\fill}
\begin{figure}[h]
\begin{center}
    \begin{minted}[fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let h_sigma = Prim.matrix k n in
  let (sigma, h), h_sigma =
    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) h_sigma
  in
  let (h_sigma, h), r_2 =
    Prim.gemm (Many 1.) (h_sigma, Many false) (h, Many true) (Many 1.) r_1
  in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
  in
  let h, new_h = Prim.copy_mat_to h h_sigma in
  let r_2, new_r = Prim.copy_mat r_2 in
  let chol_r, sol_h = Prim.posv new_r new_h in
  let chol_r, sol_data = Prim.potrs chol_r data_2 in
  let () = Prim.free_mat chol_r in
  let h_sol_h = Prim.matrix n n in
  let (h, sol_h), h_sol_h =
    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
  in
  let () = Prim.free_mat sol_h in
  let h_sol_data = Prim.matrix n (Many 1) in
  let (h, sol_data), h_sol_data =
    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
  in
  let mu, mu_copy = Prim.copy_mat mu in
  let (sigma, h_sol_data), new_mu =
    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
  in
  let () = Prim.free_mat h_sol_data in
  let h_sol_h_sigma = Prim.matrix n n in
  let (sigma, h_sol_h), h_sol_h_sigma =
    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
  in
  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
  let (sigma, h_sol_h_sigma), new_sigma =
    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
  in
  let () = Prim.free_mat h_sol_h_sigma in
  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
in
kalman
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated (at \emph{compile time})
        from the code in Figure~\ref{fig:lang_kalman}, passed through
        \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}
\end{center}
\end{figure}
\vfill

### 2018-11-05 write-up/paper/discussion.tex
Start on Soundness Theorem

Once we translated the implementation from Fortran to \lang, we attempted to
compile it and found that (to our surprise) it did not type-check. This was
because the original implementation contained incorrect aliasing, unused and
unnecessary temporaries, and did not adhere to Fortran's read/write permissions
(with respect to \texttt{intent} annotations \texttt{in}, \texttt{out} and
\texttt{inout}) all of which were now highlighted by \lang's type system.

>>>>

Once we translated the implementation from Fortran to \lang, we attempted to
compile it and found that (to our surprise) it did not type-check. This was
because the original implementation contained incorrect aliasing, unused
variables and unnecessary temporaries, and did not adhere to Fortran's
read/write permissions (with respect to \texttt{intent} annotations
\texttt{in}, \texttt{out} and \texttt{inout}) all of which were now highlighted
by \lang's type system.

### 2018-11-05 write-up/paper/formal_system.tex
Start on Soundness Theorem

The full typing rules are in Appendix \ref{subsec:static_sem}, but the key
ideas are: \textsf{Perm}, \textsf{Type}, $\Theta, \Delta, \Gamma$,
value restriction, lambda, Gen/Spc, and Fix.

@@

and operational semantics here

@@

Describe the step-indexed logical relation and its main properties

@@

State the fundamental lemma, and sketch the proof a little

>>>>

The full typing rules are in Appendix \ref{subsec:static_sem}, but the key
ideas are as follow.

A typing judgement consists of $ \Theta; \Delta; \Gamma
\vdash e : t$.

$\Theta$ is the environment that tracks which fractional permission variables
in scope. Fractional permissions (the \textsf{Perm} judgement) and types (the
\textsf{Type} judgement) are \emph{well-formed} if all of their free fractional
variables are in $\Theta$.

$\Delta$ is the environment storing non-linearly or \emph{inuitionistically}
typed variables.

$\Gamma$ is the environment storing linearly typed variables.  Note that rules
for typing $()$, booleans, integers and elements are typed with respect to an
\emph{empty} linear environment: this means no linear values are needed to
produce a value of those types.

\[
    \ottdruleTyXXUnitXXIntro{}
\]

Conversely, whenever two or more subexpressions need to be typed, they must
consume a disjoint set of linear values (pairs, let-expressions).  In the case
of if-expressions, both branches must consume the same set of linear values
(disjoint to the ones used to evaluate the condition).

\[
    \ottdruleTyXXBoolXXElim{}
\]

The \highl{Many} introduction and elimination rules are very important.
Producing !-type values may only be done if the expression inside is a
syntactic value which is not a location. This allows all safely duplicable
resources, including functions which capture non-linear resources from their
environments, but prevents producing aliases of (pointers to) arrays and
matrices. This is exactly the same as value-restriction from the world of
parametric polymorphism.

\[
    \ottdruleTyXXBangXXIntro{}
\]

Consuming a !-type value \emph{moves it} from the linear environment $\Gamma$
and \emph{into} the intuitionistic environment $\Delta$. This is exactly why
$\mathbf{let}\ !x = e_1\ \mathbf{in}\ e_2$ desugars to $\mathbf{let\ Many}\ x =
e_1\ \mathbf{in\ } \mathbf{let\ Many}\ x = \mathbf{Many}\ (\mathbf{Many}\ x)\
\mathbf{in}\ e_2$.

\[
    \ottdruleTyXXBangXXElim{}
\]

Rules \textsc{Ty\_Gen} and \textsc{Ty\_Spc} are for fractional permission
generalisation and specialisation respectively. They allow the definition and
use of functions that are polymorphic in the fractional permission index of
their results and one or more of their arguments.

\[
    \ottdruleTyXXGen{} \qquad\qquad \ottdruleTyXXSpc{}
\]

Rule \textsc{Ty\_Fix} shows how recursive functions are typed. Even though
recursive functions are fully annotated, type checking them is interesting for
two reasons: to type check the body of the fixpoint, the type of the recusive
function is in the \emph{intuitionistic} environment $\Delta$ (without this,
you would not be able to write a base case) whilst the argument and its type
are the \emph{only things in the linear environment} $\Gamma$. The latter means
that recursive functions can be type checked in an empty environment (thus be
wrapped in \highl{Many} and used zero or multiple times).

\[
    \ottdruleTyXXFix{}
\]

Lastly, types of almost all \lang\ primitives, as embedded in OCaml's type
system, are shown in Appendix \ref{subsec:primitives}, with some similar ones
(like those for binary arithmetic operators) omitted for brevity. The
main difference between the OCaml type of a primitive like \highl{gemm} and its
\lang\ counterpart is the inclusion of explicit `$\forall$'s.  So,
\highl{float bang -> ('a mat * bool bang) -> ('b mat * bool bang) -> float
bang -> z mat -> ('a mat * 'b mat) * z mat}
will correspond to \\
$!\ottkw{elt} \multimap \forall\, x.\ x \ \ottkw{mat} \ \otimes \ !\ottkw{bool}
\multimap \forall\, y.\  y \ \ottkw{mat} \ \otimes \ !\ottkw{bool} \multimap
\ !\ottkw{elt} \multimap z\ \ottkw{mat} \multimap ( x \ \ottkw{mat} \ \otimes y
\ \ottkw{mat} ) \ \otimes z\ \ottkw{mat}$

@@

The full, small-step transition relation is in Appendix \ref{subsec:dyn_sem},
but the key ideas are as follow.

Heaps ($\sigma$) are multisets containing triples of an abstract location $l$,
a fractional permission $f$ and sized matrices $m_{n,k}$. The notation $l
\mapsto_f m_{k_1, k_2}$ should be read as ``location $l$ represents $f$
ownership over matrix $m$ (of size $k_1 \times k_2$)''.  Each heap-and-expression
either steps to another heap-and-expression or a runtime error $\mathbf{err}$.
In the full grammar definition we see a definition of values and contexts in
the language.

We draw the reader's attention to the definitions relating to fractional
permissions. Specifically, unlike a lambda, the body of a $\ottkw{fun}\, f\!c
\rightarrow \_$ must be a syntactic value. The context $\ottkw{fun}\, f\!c
\rightarrow [-]$ means expressions can be reduced inside a fractional
permission generalisation. This is to emphasize that fractions are merely
\emph{compile-time constructs} and do not affect runtime behaviour. Correct
usage of fractions is enforced by the type system, so programs do not get
stuck. Fractional permissions are specialised using substitution over both the
heap and an expression (\textsc{Op\_Frac\_Perm}).
\[
    \ottdruleOpXXFracXXPerm{}
\]

Like with the static semantics, the interesting rules in the dynamic semantics
are those relating to primitives. Creating a matrix ($\ottkw{matrix}\ k_1\
k_2$) successfully (\textsc{Op\_Matrix}) requires non-negative dimensions and
returns a (fresh) location of a matrix of those dimensions, extending the heap
to reflect that $l$ represents a complete ownership over the new matrix.
\[
    \ottdruleOpXXMatrix{}
\]

Dually, \textsc{Op\_Free}, requires a location represent complete ownership
before removing it and the matrix it points to from the heap.
\[
    \ottdruleOpXXFree{}
\]

Choosing a multiset representation as opposed to a set allows for two
convenient invariants: multiplicity of a triple $l \mapsto_f m_{k_1, k_2}$ in
the heap corresponds to the number of aliases of $l$ in the expression with
type $f\ \ottkw{mat}$ and the sum of all the fractions for $l$ will always be
$1$ (for a closed, well-typed expression). With this in mind, the rules
\textsc{Op\_Share} and \textsc{Op\_Unshare\_Eq} are fairly natural.
\[
    \ottdruleOpXXShare{} \\
\]
\[
    \ottdruleOpXXUnshareXXEq{}
\]

Combining all of these features, we see that \textsc{Op\_Gemm\_Match} requires
that the location being updated ($l_3$) has complete ownership of over matrix
$m_3$ and can thus change what value it stores to $m_1 m_2 + m_3$. In
particular, this places no restriction on $l_2$ and $l_3$: they could be
$\ottkw{share}$d aliases of the same matrix. Transition rules for other
primitives (omitted) follow the same structure: $\mapsto_1$ for any locations
that are written to and $\mapsto_{f\!c}$ for anything else.
\[
    \ottdruleOpXXGemmXXMatch{}
\]

@@

First, we define an interpretation of heaps with fractional permissions in the
style of Bornat et. al~\cite{bornat} (interpreting the multiset as a partial
map from locations to the sum of all its associated fractions and a matrix) as
well as the n-fold iteration of $\rightarrow$.
\[
    \den{H}{}{\sigma} = \bigstar_{(l,f,m) \in \sigma} [ l \mapsto_f m ]
\]
where
\[
    (\varsigma_1 \star \varsigma_2)(l) \equiv
    \begin{cases}
        \varsigma_1(l) & \textrm{if } l \in \dom(\varsigma_1) \wedge l \notin \dom(\varsigma_2) \\
        \varsigma_2(l) & \textrm{if } l \in \dom(\varsigma_2) \wedge l \notin \dom(\varsigma_1) \\
        (f_1 + f_2, m) & \textrm{if } (f_1, m) = \varsigma_1(l) \wedge (f_2, m) = \varsigma_2(l) \wedge f_1 + f_2 \leq 1 \\
        \textrm{undefined} & \textrm{otherwise}
    \end{cases}
\]

We then define a step-indexed logical relation in the style of Morrisett et.
al~\cite{morrisett}. $(\varsigma, v) \in \V{k}{t}$ means it takes a heap with
exactly $\varsigma$ resources to produce a value $v$ of type $t$ in at most $k$
steps. So, something like a $\ottkw{unit}$ or a $!t$ need no resources, whereas
a $f\, \ottkw{mat}$ needs exactly $f$ ownership of a some matrix and a pair
needs a $\star$ combination of the heaps required for each component.
\begin{align*}
  \V{k}{ \Unit } &= \{ (\empH, \ast) \} \\
  \V{k}{ f \, \Mat } &= \{ (\{ l \mapsto_{2^{-f}} \_ \} , l) \} \\
  \V{k}{ \Bang t } &= \{ (\empH, \Many\, v) \mid (\empH, v) \in \V{k}{t} \} \\
  \V{k}{ t_1 \otimes t_2 } &= \{ (\varsigma_1 \star \varsigma_2, ( v_1, v_2 )) \mid (\varsigma_1, v_1) \in \V{k}{t_1} \wedge (\varsigma_2, v_2) \in \V{k}{t_2} \}
\end{align*}

The definition of $\V{k}{\forall f\!c.\ t}$ says a value and heap
must be the same regardless of what fraction is substituted into both; the
$k-1$ is to take into account fraction specialisation takes ones step
(\textsc{Op\_Spc}).
\[
    \V{k}{ \forall f\!c.\  t } = \{ (\varsigma, \ottkw{fun}\, f\!c \rightarrow \, v) \mid \forall f.\ (\varsigma [ f\!c / f ], v [ f\!c / f ]) \in \V{k-1}{ t [ f\!c / f ] } \}
\]

To understand the definition of $\V{k}{t' \multimap t}$, we must first look at
$\C{k}{t}$, the computational interpretation of types. Intuitively, it is a
combination of a frame rule on heaps (no interference), type-preservation and
termination (in $j < k$ steps) to either an error or a heap-and-expression,
with the further condition that if the expression is a syntactic value then it
is also one semantically.
\begin{align*}
    \C{k}{ t } &= \{ (\varsigma_s, e_s) \mid \forall \, j < k, \sigma_r.\ \varsigma_s \star \varsigma_r \textrm{ defined } \Rightarrow \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \ottkw{err}\ \vee \exists \sigma_f, e_f.\\
               & \qquad \qquad \langle \sigma_s + \sigma_r, e_s \rangle \rightarrow^j \langle \sigma_f + \sigma_r, e_f \rangle \wedge ( e_f \textrm { is a value } \Rightarrow ( \varsigma_f \star \varsigma_r, e_f ) \in \V{k-j}{t} ) \}
\end{align*}

In this light, $\V{k}{t' \multimap t}$ simply says
that $v$ is a function and that evaluating the application of it to any
argument (of the correct type, requiring its own set of resources, bounded by
$k$ steps) satisfies all the aforementioned properties.
\begin{align*}
    \V{k}{ t' \multimap t } &= \{ (\varsigma_v, v ) \mid ( v \equiv \ottkw{fun}\, x : t' \rightarrow e \vee v \equiv \ottkw{fix} (g, x : t' , e : t) ) \, \wedge\\
                            & \qquad \qquad \forall j \leq k, (\varsigma_{v'}, v') \in \V{j}{ t' }.\ \varsigma_v \star \varsigma_v' \textrm{ defined } \Rightarrow (\varsigma_v \star \varsigma_v', v\, v') \in \C{j}{t} \}
\end{align*}

The interpretation of typing environments $\Delta$ and $\Gamma$ are with
respect to an arbitrary substitution of fractional permissions $\theta$. Note
that only the interpretation of $\Gamma$ involves a non-empty heap.
\begin{align*}
    \den{I}{k}{ \Delta, x : t } \theta &= \{ \delta[x \mapsto v_x ] \mid \delta \in \den{I}{k}{\Delta}\theta \wedge (\empH, v_x) \in \V{k}{\theta(t)} \} \\
    \den{L}{k}{ \Gamma, x : t } \theta &= \{ (\varsigma \star \varsigma_x, \gamma[x \mapsto v_x ]) \mid (\varsigma, \gamma) \in \den{L}{k}{\Gamma}\theta \wedge (\varsigma_x, v_x) \in \V{k}{\theta(t)} \}
\end{align*}

And so the final semantic interpretation of a typing judgement simply
quantifies over all possible fractional permission substitutions $\theta$,
linear value substitutions $\gamma$, intuitionistic value substitutions
$\delta$ and heaps $\sigma$.  Note that, $\varsigma \equiv \den{H}{}{\theta(\sigma)}$.
\begin{align*}
\den{}{k}{ \Theta; \Delta ; \Gamma \vdash e : t } &= \forall \theta, \delta, \gamma, \sigma.\ \Theta = \dom(\theta) \wedge (\varsigma, \gamma) \in \den{L}{k}{ \Gamma }\theta \wedge \delta \in \den{I}{k}{ \Delta }\theta \Rightarrow \\
                                                     & \qquad \qquad (\varsigma, \theta(\delta(\gamma(e)))) \in \C{k}{ \theta(t) }
\end{align*}

@@

\[
    \forall \Theta, \Delta, \Gamma, e, t.\ \Theta; \Delta ; \Gamma \vdash e : t \Rightarrow
    \forall k.\ \den{}{k}{ \Theta; \Delta ; \Gamma \vdash e : t }
\]

To prove the above theorem, we need several lemmas; the interesting ones are:
\ref{frame} (a proof of the frame rule), \ref{subsetKJ} (monotonicity for the
step-index), \ref{restriction} (splitting up environments corresponds to
splitting up heaps) and \ref{fracPermSub} (a heap-and-expression take the same
steps of evaluation under any substitution of their free fractional
permissions).

The proof proceeds by induction on the typing judgement.
